---
code-fold: true
---

```{r}
#| label: set-up
#| message: false
#| echo: false
library(tidyverse)
library(webexercises)
add_to_quarto()
```

# Exercises

## Getting started

### Exploring RStudio

1 . Find the keyboard shortcuts menu:

Tools \> Keyboard shortcuts help

2.  Change the appearance to something different.

### Installing packages and using functions

3. Loading packages

```{r}
#| label: cowsay 
#| echo: false
opts <- c("install.package", 
            "install.packages", 
            answer = "library", 
            "libraries")

q1 <- mcq(opts)

cat("What function loads a package that is already on your computer?", q1)
```

### Creating folders

Assuming we have created a project called `library-r`:

1.  Create a folder called `R` and folder called `outputs` in your project folder

## Palmer penguins

Load the Palmer Penguins library if it's not already loaded.

### Palmer penguins dataset

1.  Find out about the `penguins` dataset:

-   what is it?
-   and what data types does it contain?

### Vectors and assignment

```{r} 
#| label: test-q
cat("Type the assignment operator ", fitb('<-'))
```

1.  Create a character vector of your name and assign it to an object called my_name

    ```{r}
    #| label: my-name
    #| eval: false
    my_name <- "Alistair"
    ```

2.  Pass my_name to cowsay as an argument. Chose whatever animal you wish

    ```{r}
    #| label: monkey_name
    #| eval: false

    say(my_name, by = "monkey")
    ```

3.  Create a sequence of numbers from 1 to 10 and assign it to an object called my_seq

    ```{r}
    #| label: my-seq
    #| eval: false
    my_seq <- seq(1:10)
    ```

### Data frames/tibble

1.  Make a character vector of three names
2.  Make a numeric vector of three numbers
3.  Make a factor vector of three fruit
4.  Combine into a data frame.

```{r}
#| label: df-example
# For example
char_vec <- c("James","Hannah","Matt")
num_vec <- c(5,1792,23)
factor_vec <- factor(x = c("oranges","apples","grapes"))

# Combine into data.frame, the names of the variables will
# come from the objects.
df <- data.frame(char_vec,
                 num_vec,
                 factor_vec)
```

## dplyr

All these exercises use `dplyr` from the tidyverse, but in addition to reading the book, you may have to Google or ask an AI assistance for help with some of these exercises.

### Importing and inspecting data

Exercise 1: Import the data contained in csv file to an object called `books` from: <https://raw.githubusercontent.com/ab604/library-r/main/data/books-2024-04-30.csv>

::: callout-tip
You can read data directly from paths to on-line datasets with `read_csv()` as you would for locally stored `csv` files.
:::

```{r}
#| label: import-books
#| message: false
# Exercise 1: Import the data contained in csv file to an object called `books`
books <- read_csv("https://raw.githubusercontent.com/ab604/library-r/main/data/books-2024-04-30.csv")
```
