---
title: "R for Librarians Week 2 Exercises"
author: "Alistair Bailey"
date: last-modified
date-format: "[Last Updated on] YYYY-MM-DD"
format: 
    html:
      code-fold: true
      code-link: true
    pdf:
      toc: true
      number-sections: true
      colorlinks: true
knitr:
  opts_chunk:
    results: hide
execute: 
  eval: true
  cache: true
---

```{r}
#| label: set-up
#| message: false
#| echo: false
library(tidyverse)
library(palmerpenguins)
library(glue)
```

## Warm-up

1.  Create a Quarto markdown project called `r-week-02` and edit the default document it opens to use for our work today and save it a new folder called `R`.
2.  From the console pane, install the `palmerpenguins` package (feel free to Google or look back at last weeks workbook to see how to do this).
3.  Create a code chunk in the editor pane to load the `palmerpenguins` package. Label the chunk `set-up`.

## Exercises

### Vectors and assignment

1.  Create a character vector of your name and assign it to an object called my_name

    ```{r}
    #| label: my-name
    #| eval: false
    my_name <- "Alistair"
    ```

2.  Pass my_name to cowsay as an argument. Chose whatever animal you wish

    ```{r}
    #| label: my_name
    #| eval: false

    say(my_name, by = "monkey")
    ```

3.  Create a sequence of numbers from 1 to 10 and assign it to an object called my_seq

    ```{r}
    #| label: my-seq
    #| eval: false
    my_seq <- seq(1:10)
    ```

4.  Compare the `dplyr` package `glimpse()` function and the base R `str()` function for summarizing the contents `penguins` data.

```{r}
#| label: summarise-functions
glimpse(penguins)
str(penguins)
```
