---
title: "R for Librarians Introduction Exercises"
author: "Alistair Bailey"
date: last-modified
date-format: "[Last Updated on] YYYY-MM-DD"
format: 
    html:
      code-fold: true
      code-link: true
    pdf:
      toc: true
      number-sections: true
      colorlinks: true
knitr:
  opts_chunk:
    results: hide
execute: 
  eval: true
  cache: true
---

```{r}
#| label: set-up
#| message: false
#| echo: false
library(tidyverse)
```

### Exploring RStudio

1 . Find the keyboard shortcuts menu:

Tools \> Keyboard shortcuts help

2.  Change the appearance to something different.

### Installing packages and using functions

3 .Install cowsay and make a cat say "U openz ur research to me, I openz my hear to U"

```{r}
#| label: cowsay 
#| eval: false
install.packages("cowsay")
library(cowsay)

# Make a cat declare their love of open access publishing
say("U openz ur research to me, I openz my hear to U", by = "cat")
```

4.  Roundhouse

```{r}
#| label: roundhouse
#| eval: false
install.packages("roundhouse")
library(roundhouse)
random_fact()
kick()
```

5.  Open the help files for:

-   function `animals()` from the `cowsay` package
-   function `punch()` from the `roundhouse` pacakage

```{r}
#| label: help-pages
#| eval: false
?cowsay::animals
?roundhouse::punch
```

### Creating folders

Assuming we have created a project called `library-r`:

5.  Create a folder called `R` and folder called `outputs` in your project folder

## Palmer penguins

Load the Palmer Penguins library if it's not already loaded.

### Palmer penguins dataset

1.  Find out about the `penguins` dataset:

-   what is it?
-   and what data types does it contain?

### Vectors and assignment

1.  Create a character vector of your name and assign it to an object called my_name

    ```{r}
    #| label: my-name
    #| eval: false
    my_name <- "Alistair"
    ```

2.  Pass my_name to cowsay as an argument. Chose whatever animal you wish

    ```{r}
    #| label: my_name
    #| eval: false

    say(my_name, by = "monkey")
    ```

3.  Create a sequence of numbers from 1 to 10 and assign it to an object called my_seq

    ```{r}
    #| label: my-seq
    #| eval: false
    my_seq <- seq(1:10)
    ```

### Data frames/tibble

1.  Make a character vector of three names
2.  Make a numeric vector of three numbers
3.  Make a factor vector of three fruit
4.  Combine into a data frame.
