---
title: "R for Librarians Week 3 Warm-up Exercises"
author: "Alistair Bailey"
date: last-modified
date-format: "[Last Updated on] YYYY-MM-DD"
format: 
    html:
      code-fold: true
      code-link: true
    pdf:
      toc: true
      number-sections: true
      colorlinks: true
knitr:
  opts_chunk:
    results: hide
execute: 
  eval: true
  cache: true
---

```{r}
#| label: set-up
#| message: false
#| echo: false
library(tidyverse)
library(palmerpenguins)
```

## Warm-up

1.  Create a Quarto markdown project called `r-week-03` and edit the default document it opens to use for our work today and save it.

2.  Create a code chunk in the editor pane to load the `tidyverse` and `palmerpenguins` packages. Label the chunk `set-up`.

```{r}
#| label: set-up-show
#| eval: false
#| message: false
library(tidyverse)
library(palmerpenguins)
```

3. Create a code chunk and pipe the `penguins` object to the function `glimpse()` and execute the code.

```{r}
#| label: glimpse
#| eval: false
penguins |> 
  glimpse()
```


4. Write and run some code that filters the `penguins` data for flipper less than to 180 mm. 

```{r}
#| label: filter-bill
#| eval: false
penguins |> 
  filter(flipper_length_mm < 180)
```

