---
title: "R for Librarians `dplyr` Exercises"
author: "Alistair Bailey"
date: last-modified
date-format: "[Last Updated on] YYYY-MM-DD"
format: 
    html:
      code-fold: true
      code-link: true
      embed-resources: true
    # pdf:
    #   toc: true
    #   number-sections: true
    #   colorlinks: true
knitr:
  opts_chunk:
    results: hide
execute: 
  eval: true
  cache: true
---

Use the `tidyr` data: https://tidyr.tidyverse.org/reference/index.html#data and
https://nycflights13.tidyverse.org/ data

```{r}
#| label: set-up
#| message: false
library(tidyverse)
```

```{r}
#| label: relig-income
#| message: false
relig_income <- relig_income

relig_income
```


### Exercise 1: Pivot Longer

```{r}
relig_income |> 
  pivot_longer(
    cols = -religion, 
    names_to = "income", 
    values_to = "count"
  )
```

Your turn! Use the `pivot_longer()` function to transform the `penguins` dataset from wide to long format. Then, try to reconstruct the original dataset from the long format.

### Exercise 2: Pivot Wide to Long

```{r}
penguins |>
  pivot_longer(-species, names_to = "trait", values_to = "value") |>
  pivot_wider(id_cols = species, names_from = "trait", values_from = "value") |>
  head()
```

In this exercise, you'll use the `pivot_longer()` and `pivot_wider()` functions to transform the `penguins` dataset from long to wide format to long to wide format. Can you figure out how to get back to the original long format from the wide format?

### Exercise 3: Summarise and Pivot

```{r}
penguins |>
  group_by(species) |>
  summarise(avg_bill_length_mm = mean(culmen_length_mm)) |>
  pivot_wider(names_from = species, values_from = avg_bill_length_mm) |>
  head()
```

In this exercise, you'll create a pivot table using the `group_by()` and `summarise()` functions. Then, use the `pivot_wider()` function to transform the data from long to wide format. Finally, reflect on how the resulting table could be useful in your analysis.
