---
title: "Read excel practice"
---

```{r}
library(tidyverse)
library(janitor)
library(readxl)
```

## Read in

> "...a fortnightly check that Stephie carries out for REF and Funder Compliance, the Export Control Team periodically asks us for information about particular staff/institutions/countries and we can use this as it has been sense checked as UoS, trying to search in the Affiliations can be very time consuming..."

```{r}
dat <- read_xlsx("../data/V1 New OA Compliance Master Sheet.xlsx", sheet = "Master (New data)") 
```

```{r}
dat |> filter(`UoS Corresponding Author` == "Yes") |> tally()

dt <- dat |> #select(Affiliations) |> 
  separate_wider_delim(cols = Affiliations, delim = ";", names_sep = "_", 
                       too_few = "align_start")

dt_long <- dt |> pivot_longer(14:883, names_to = "Affiliation_No", values_to = "Department", values_drop_na = TRUE)


dt_uos <- dt_long |> filter(str_detect(Department, "South*ampton"))

dt_uos |> distinct(Title) |> tally()
dt_long |> distinct(Title) |> tally()

dt_uos |> group_by(Department) |> count() |> view()

dt_uos |> filter(str_detect(Department,"MRC Lifecourse Epid")) |> distinct(Title) |> tally()

dt_uos |> summarise_all(list(~ sum(is.na(.))))

dt_long |> count(Department, sort = T)

dt_all <- dat |> left_join(dt_uos)

dat |> anti_join(dt_uos) |> view()

# Count the missing values
dat |> summarise_all(list(~ sum(is.na(.)))) |> 
  pivot_longer(1:35, names_to = "variable", values_to = "NAs") |> view()
```

To-do; create synthetic data, sense checks and counting NAs, looking for spelling inconsistencies.
